Polynomial Time and Full Dependent Types

### 2023-07-10 paper/paper.tex
[ paper ] lots more writing

\subsection{No Recursion, Only Case Analysis}

@@

\subsection{Diamond Trading with LFPL}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Quantitative Type Theory}
\label{sec:qtt}

We have now seen two systems for controlling resource usage by means
of linear typing. We now look to extend these systems to include
dependent types. Our motivation is threefold:
\begin{enumerate}
\item We would like to have a means for reasoning about the behaviour
  of the programs we write in our systems. FIXME: why dependent types
  then? Why not a program logic?
\item In case of the explicitly resource accounted system, simple
  typing is limited in the reasoning the programmer can perform about
  the resources required by a particular program. While we could in
  principle
\item As well as constructing proofs \emph{about} complexity
  constrained programs, we would also be able to write constructive
  proofs that are themselves of constrained complexity.
\end{enumerate}

To integrate linear typing for controlling time complexity with
dependent types we use \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt}. In this section, we review the definition of QTT
and how we have adapted it to the polytime systems we saw in the
previous section.

@@

\bob{Explain the judgements and general idea of QTT, and the admissibility of $0$-ing}

\subsubsection{Contexts, variables, and conversion}
\begin{mathpar}
  \inferrule* [right=Ctxt-Emp]
  { }
  {\epsilon~\isctxt}

  \inferrule* [right=Ctxt-Ext]
  {\Gamma~\isctxt \\ 0\Gamma \vdash S~\istype}
  {\Gamma, x \stackrel\rho: S~\isctxt}
\end{mathpar}

\begin{mathpar}
  \inferrule* [right=Tm-Var]
  {0\Gamma, x \stackrel\sigma: S, 0\Gamma'~\isctxt}
  {0\Gamma, x \stackrel\sigma: S, 0\Gamma' \vdash x \stackrel\sigma: S}

  \inferrule* [right=Tm-Conv]
  {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
  {\Gamma \vdash M \stackrel\sigma: T}
\end{mathpar}

\subsubsection{$\Pi$- and $\Sigma$-types}
\begin{mathpar}
  \inferrule* [right=$\Pi$-type]
  {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
  {0\Gamma \vdash (x \stackrel\rho: S) \to T~\istype}

  \inferrule* [right=$\Pi$-Intro]
  {\Gamma, x \stackrel{\sigma\rho}: S \vdash M \stackrel\sigma: T}
  {\Gamma \vdash \lambda x.M \stackrel\sigma: (x \stackrel\rho: S) \to T}

  \inferrule* [right=$\Pi$-Elim]
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    FIXME}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}

\bob{$\Sigma$-types, from the LICS paper}

\subsubsection{The Identity Type}
\begin{mathpar}
  \inferrule* [right=Eq-type]
  {0\Gamma \vdash S~\istype \\
    0\Gamma \vdash M \stackrel0: S \\
    0\Gamma \vdash N \stackrel0: S}
  {0\Gamma \vdash M =_S N~\istype}

  \inferrule* [right=Eq-Intro]
  {\Gamma \vdash M \stackrel\sigma: S}
  {\Gamma \vdash \conRefl(M) \stackrel\sigma: M =_S M}
\end{mathpar}
\bob{Equality reflection}

\subsubsection{Universe}
\bob{Copy from LICS paper}

\subsubsection{Datatypes} QTT, as we have presented it so far, has no
interesting base types to perform computation on. Following our
presentation of the simply typed linear systems in
\autoref{sec:lin-intro}, we add two kinds of datatype to QTT. First,
\autoref{sec:noniter-qtt}, we add non-iterable datatypes that allow
construction and case analysis, but no recursion. In
\autoref{sec:cons-free-qtt} we describe how to extend QTT to be a
dependently typed adaptation of the Cons-free system of
\autoref{sec:cons-free-intro}. In \autoref{sec:lfpl-qtt}, we apply the
same treatment to the LFPL-style system.

@@

\subsubsection{Booleans} The boolean type was already described in
\cite{atkey18qtt}, but we repeat it here for completeness. Booleans
offer no possibility for iteration in any case, but it is useful to
see how their typing rules handle resource distribution before moving
to more complex types.
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {\Gamma \vdash \BoolTy~\istype}

@@

\bob{Reword this paragraph in light of whatever is said in the
  previous section.} The datatypes of the previous section still only
allow us to write programs in the $\sigma = 1$ fragment that are
constant time in the size of their input. If we are handed a list of
an unknown length, we can only explore it to a fixed depth determined
statically. As we saw above, to write programs that perform work
proportional to the length of their input, we need some form of
iterable datatype. In both our cons-free and LFPL-style systems, we
use a natural number datatype.

@@

The reader is invited to compare this dependently typed rule with the
simply typed linear version in \autoref{sec:cons-free-intro}. Removing
the $0$-annotated parts of the rule, and the type dependency, yield
the exact same rule. Conversely, when $\sigma = 0$, this rule is
identical (up to $0$-annotations) to the usual dependently typed
recursor for natural numbers, and so we can use it in the types to
prove properties of programs just as we do in normal type theory. We
will see in \autoref{sec:cons-free-sound} that this rule is realisable
by polynomial-time amortised computation, and so is sound for
polynomial time. \bob{And we do some examples?}

@@

\subsection{The Cons-Free System}

@@

\subsection{The LFPL System}

>>>>

\subsection{No Recursion, Only Case Analysis}
\label{sec:linear-case-analysis}

@@

\subsection{Diamond Trading with LFPL}
\label{sec:lfpl-intro}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Polytime Quantitative Type Theory}
\label{sec:qtt}

We have now seen the \emph{Cons-free} and \emph{LFPL} systems for
capturing polytime by means of linear typing and restricted
iteration. We now look to extend these systems to include dependent
types.

To integrate linear typing for controlling time complexity with
dependent types we use \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt}. We now review the definition of QTT and describe
how we have adapted it to the polytime systems we saw in the previous
section.

@@

Integrating linear and dependent types is not straightforward due to
the conflict between the linear typing discipline regarding presence
of a variable as only bestowing the right to use it once, and the
dependent typing regime that uses variables both in types (for
specification purposes) and in terms (for computational purposes),
syntactically yield multiple uses of the same variable.

QTT is a system that resolves this conflict by recording usage of
variables with annotations from a semiring. It sits in the general
area of systems that use semiring annotatons to measure resource usage
\cite{BLAH}. The key feature of QTT, originally due to
\cite{mcbride16}, is that usage of variables in types counts for
$0$-usage in terms of the semiring used. This allows us to use normal
type theory as a specification language, while also enjoying the
benefits of linear typing for programs. The term typing judgement of
QTT has the following form:
\begin{displaymath}
  x_1 \stackrel{\rho_1}: S_1, \dots, x_n \stackrel{\rho_n}: S_n \vdash M \stackrel\sigma: T
\end{displaymath}
where the annotations $\rho_i$ are all from the semiring being
used. The annotation $\sigma$ is either $0$ or $1$, indicating whether
we are in the erased ($\sigma = 0$) fragment, where all the normal
rules of type theory apply, or the in the non-erased (``present'',
$\sigma = 1$) fragment, where a restricted typing discipline applies.
As we shall see below in the cases of $\Sigma$-types, iterable types,
and LFPL's $\Diamond$ type, the separation of QTT into these two
fragments allows an expressive combination of reasoning using full
type theory with the benefits of linear typing.

In the remainder of this sub-section, we describe the core of QTT. As
well as the term typing judgement given above, QTT also has judgements
for well-formed contexts ($\Gamma~\isctxt$) and types
($\Gamma \vdash T~\istype$), and definition equality of types and
terms ($\Gamma \vdash S \equiv T~\istype$ and
$\Gamma \vdash M \equiv N \stackrel\sigma: S$). It is an invariant of
the system that types are always well-formed in a context with all
annotations $0$, i.e., $\Gamma \vdash S~\istype$ implies
$0\Gamma = \Gamma$. An important admissible rule of the system, along
with weakening and substitution, is that of $0$-ing:
\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M \stackrel1: S}
  {0\Gamma \vdash M \stackrel0: S}
\end{displaymath}
This rule allows us to take any term $M$ in the $\sigma = 1$ fragment
and treat it as if it were in the $\sigma = 0$ fragment, and hence use
it for specification purposes in types. As we add novel rules to QTT
in the following sections, we will be careful to maintain the
admissibility of this rule.

\subsubsection{Natural-number usages}

We use an instantiation of QTT with the natural number semiring, with
the usual semiring structure of addition and multiplication. In a mild
generalisation of the original presentation of QTT, we also allow
sub-usaging via the \emph{reverse} ordering on the naturals. That is,
if a variable is marked as usage $n$ and $m \geq n$, then we can also
regard it as usage $m$. This makes the system more like affine linear
logic, since $m \leq 0$ for all $m$, matching the system in
\autoref{sec:linear-intro}. We do \emph{not} have an unrestricted
usage $\omega$, since this would allow the possibility of unrestricted
duplication, and hence violate our polytime soundness properties.

\subsubsection{Contexts, variables, and conversion}

As we saw above, contexts in QTT are comprised of variable
$\stackrel\rho:$ type triples, where $\rho$ is a natural number
indicating how many times the variable $x$ is available for use in a
$\sigma = 1$ term. There are two operations on raw contexts: scaling
$\pi\Gamma$, which multiplies each $\rho$ in $\Gamma$ by $\pi$, and
addition $\Gamma_1 + \Gamma_2$, which adds two contexts' usage
annotations assuming that the lengths and types are equal.  The basic
usage-annotation discipline of QTT is demonstrated by the context
formation and variable rules:
\begin{mathpar}
  \inferrule*
  { }
  {\epsilon~\isctxt}

  \inferrule*
  {\Gamma~\isctxt \\ 0\Gamma \vdash S~\istype}
  {\Gamma, x \stackrel\rho: S~\isctxt}

  \inferrule*
  {0\Gamma, x \stackrel\sigma: S, 0\Gamma'~\isctxt}
  {0\Gamma, x \stackrel\sigma: S, 0\Gamma' \vdash x \stackrel\sigma: S}
\end{mathpar}
As with most dependent type theories, contexts are built inductively
from the empty context $\epsilon$ and extension of a context by a
variable with a type that is well-formed in the preceding
context. Usage annotations $\rho$ on variables are arbitrary, and
types are always judged in a $0$-annotated context. The variable rule
marks unused variables as usage $0$ and the selected variable with
usage $\sigma$.

% \begin{mathpar}

%   \inferrule* [right=Tm-Conv]
%   {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
%   {\Gamma \vdash M \stackrel\sigma: T}
% \end{mathpar}

\subsubsection{$\Pi$- and $\Sigma$-types}

QTT's $\Pi$-types have the form $(x \stackrel\rho: S) \to T$,
indicating functions that, in the $\sigma = 1$ fragment, use their
arguments $\rho$-many times. The formation, introduction and
elimination rules are similar to the standard ones, except for the
addition of usage annotations:
\begin{mathpar}
  \inferrule*
  {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
  {0\Gamma \vdash (x \stackrel\rho: S) \to T~\istype}

  \inferrule*
  {\Gamma, x \stackrel{\sigma\rho}: S \vdash M \stackrel\sigma: T}
  {\Gamma \vdash \lambda x.M \stackrel\sigma: (x \stackrel\rho: S) \to T}

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}
The side conditions on the elimination rule state that (i) both
$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
either the function does not use its argument, or we are in the
$\sigma = 0$ fragment and everything is being erased.

$\Sigma$-types are a little more involved, and demonstrate the
flexibility in QTT in allowing additional power in the $\sigma = 0$
fragment, where we do not need to care about
implementability. Formation and introduction are given by the rules:
\begin{mathpar}
  \inferrule*
  {0\Gamma \vdash S \\ 0\Gamma, x \stackrel0: S \vdash T}
  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T}

  \inferrule*
  {\Gamma_1 \vdash M \stackrel{\sigma'}: S \\
    \Gamma_2 \vdash N \stackrel\sigma: T[M/x] \\\\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0)}
  {\pi\Gamma_1 + \Gamma_2 \vdash (M, N) \stackrel\sigma: (x \stackrel\pi: S) \otimes T}
\end{mathpar}
As with $\Pi$-typed, the first component of a $\Sigma$ type is
annotated with a usage for how many times it can be used, and this is
respected by the introduction rule. Elimination of $\Sigma$-types
depends on whether we are in the $\sigma = 0$ fragment or not. In the
$\sigma = 0$ fragment, we are free to disregard usage restrictions,
and use projections as normal:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{fst}(M) \stackrel0: S}

  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
\end{mathpar}
Hence, $\Sigma$-types act as normal in the $\sigma = 0$ fragment, and
we can use them as for normal type-theoretic constructions.  In the
$\sigma = 1$ fragment, we must take into account the resource content
of objects and use a pattern matching construct; the dependently typed
analogue of the $\otimes$-eliminator in \autoref{sec:lineartype}:
\begin{mathpar}
  \inferrule*
  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
\end{mathpar}

\subsubsection{The Identity Type}

QTT also supports an extensional equality type with equality reflection:
\begin{mathpar}
  \inferrule*
  {0\Gamma \vdash S~\istype \\
    0\Gamma \vdash M \stackrel0: S \\
    0\Gamma \vdash N \stackrel0: S}
  {0\Gamma \vdash M =_S N~\istype}

  \inferrule*
  {\Gamma \vdash M \stackrel\sigma: S}
  {\Gamma \vdash \conRefl(M) \stackrel\sigma: M =_S M}

  \inferrule*
  {\Gamma \vdash N \stackrel\sigma: M_1 =_S M_2}
  {\Gamma \vdash M_1 \equiv M_2 \stackrel\sigma: S}
\end{mathpar}
The equality type also has an $\eta$ rule demonstrating $\conRefl(M)$
as the only proof of equality \cite{hoffmann99dependent}.

\subsubsection{Universe} QTT can also be equipped in universe types,
just as in standard type theory. We refer to the rules presented in
\cite{atkey18qtt}.

\subsubsection{Datatypes} QTT, as we have presented it so far, has no
interesting base types to perform computation on. Following our
presentation of the simply typed linear systems in
\autoref{sec:linear-intro}, we add two kinds of datatype to QTT. First,
\autoref{sec:noniter-qtt}, we add non-iterable datatypes that allow
construction and case analysis, but no recursion. In
\autoref{sec:cons-free-qtt} we describe how to extend QTT to be a
dependently typed adaptation of the Cons-free system of
\autoref{sec:cons-free-intro}. In \autoref{sec:lfpl-qtt}, we apply the
same treatment to the LFPL-style system.

@@

\subsubsection{Booleans} The boolean type for QTT was described in
\cite{atkey18qtt}, but we repeat it here for completeness. Booleans
offer no possibility for iteration, but it is useful to see how the
QTT versions of the rules extend the simply typed rules from
\autoref{sec:linear-case-analysis} before moving to more complex
types.
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {\Gamma \vdash \BoolTy~\istype}

@@

The datatypes of the previous section still only allow us to write
programs in the $\sigma = 1$ fragment that are constant time in the
size of their input. As with the simply typed linear system, if we are
handed a list of an unknown length, we can only explore it to a fixed
depth determined statically. To write programs that do work
proportional to the size of their input, we need some form of iterable
datatype. In both our Cons-free and LFPL-style QTT systems, we use a
natural number datatype.

@@

The reader is invited to compare this dependently typed rule with the
simply typed linear version in \autoref{sec:cons-free-intro}. Removing
the $0$-annotated parts of the rule, and the type dependency, yield
the exact same rule. Conversely, when $\sigma = 0$, this rule is
identical (up to $0$-annotations) to the usual dependently typed
recursor for natural numbers, and so we can use it in the types to
prove properties of programs just as we do in normal type theory. We
will see in \autoref{sec:cons-free-sound} that this rule is realisable
by polynomial-time amortised computation, and so is sound for
polynomial time.

@@

\subsection{The Cons-Free System}
\label{sec:cons-free-sound}

@@

\subsection{The LFPL System}
\label{sec:lfpl-sound}
